---
title: "Parse CPTAC methylation array"
output: html_notebook
---
Parse the beta values of the probes in the EPIC DNA methylation array. Then filter out some low confident probes and subset the samples to be discovery cohort only.

```{r, message=FALSE}
library(feather)
library(SummarizedExperiment)
library(tidyverse)
```


Read the beta values per probe
```{r}
meth_tbl <- read_feather('../processed_data/cptac3_gbm_epic_meth_array.feather')
```

Read EPIC methylation array hg38 manifest and annotaiton from <https://zwdzwd.github.io/InfiniumAnnotation>
```{r}
# Details about the probe design
EPIC_manifest <- readRDS('../external_data/EPIC.hg38.manifest.rds')
# Annotation
EPIC_anno <- readRDS('../external_data/EPIC.hg38.manifest.gencode.v22.rds')
```

There are some probes not in the EPIC manifest file
```{r}
probes_not_in_manifest <- setdiff(meth_tbl$probe_id, names(EPIC_manifest))
length(probes_not_in_manifest)
```

```{r}
probes_not_in_manifest[1:5]
```

But those missing probes are actually available in the `IlluminaHumanMethylationEPICanno.ilm10b2.hg19`'s annotation. The reason they are missing remain unknown. To investigate this issues, load the illumina's manifest and annotaiton in hg19 by:

```r
library(IlluminaHumanMethylationEPICanno.ilm10b2.hg19)
EPIC_hg19_anno <- getAnnotation(IlluminaHumanMethylationEPICanno.ilm10b2.hg19)
EPIC_hg19_anno[probes_not_in_manifest, ]
```

## Add sample annotation
```{r}
sample_anno <- read_csv('../../annotations/Batch1through8_sample_attributes.csv.gz')
case_list <- read_tsv('../../201904_locate_adhoc_data/tracked_results/CPTAC3_GBM_all_cases.tsv')
```

```{r}
coldata_df <- sample_anno %>%
    filter(SAMPLE_ID %in% colnames(meth_tbl)[-1]) %>%
    mutate(
        is_tumor = IS_TUMOR == 'Y',
        cancer_type = 'GBM',
        primary_tumor_location = factor(PRIMARY_TUMOR_LOCATION)
    ) %>%
    select(
        sample=SAMPLE_ID,
        case=SUBJECT_ID,
        cancer_type,
        is_tumor,
        primary_tumor_location
    ) %>%
    filter(case %in% case_list$case) %>%
    arrange(case) %>%
    column_to_rownames('sample') 
coldata_df %>% head()
```

```{r}
nrow(coldata_df)
```


## Use the masked probes
Number of probes before filtering
```{r}
nrow(meth_tbl)
```

Select only probes not masked
```{r}
selected_probes <- intersect(meth_tbl$probe_id, EPIC_manifest %>% subset(!MASK_general) %>% names())
length(selected_probes)
```

Convert the methylation array to a matrix
```{r}
meth_mat <- meth_tbl %>%
    filter(probe_id %in% selected_probes) %>%
    column_to_rownames('probe_id') %>% 
    as.matrix()

meth_mat <- meth_mat[, rownames(coldata_df)]

dim(meth_mat)
```

```{r}
selected_anno_gr <- EPIC_anno[selected_probes, ]
```


## Create a SummarizedExperiment object
```{r}
meth_se <- SummarizedExperiment(
    rowRanges = selected_anno_gr,
    colData = coldata_df,
    assays = list(beta_val=meth_mat),
    metadata = list(
        cohort = 'CPTAC3 GBM adhoc cohort (58 cases)',
        description = 'Beta values of DNA methylation probes from Illumina EPIC microarry',
        pipeline = str_c(
            'Based on the Bioconductor workflow methylationArrayAnalysis. ',
            'Pipeline details at https://github.com/ding-lab/cptac_methylation. ',
            'Only select probes with MASK_general == FALSE from the hg38 manifest file ',
            'at http://zwdzwd.io/InfiniumAnnotation/20180909/EPIC/EPIC.hg38.manifest.rds. ',
            'More information at https://zwdzwd.github.io/InfiniumAnnotation (DOI: 10.1093/nar/gkw967)'
        ),
        annotation = str_c(
            'Based on GENCODE v22 ',
            'from http://zwdzwd.io/InfiniumAnnotation/20180909/EPIC/EPIC.hg38.manifest.gencode.v22.rds'
        ),
        contact = 'Wen-Wei Liang <liang.w@wustl.edu>, Liang-Bo Wang <liang-bo.wang@wustl.edu>'
    )
)
```

```{r}
saveRDS(meth_se, '../processed_data/gbm_dna_meth.rds')
```

```{r}
coldata_df %>%
    pull(case) %>%
    write_lines('../processed_data/cases_with_dna_meth.list')
```


