---
title: "Parse CPTAC methylation array"
output: html_notebook
---
Parse the beta values of the probes in the EPIC DNA methylation array. Then filter out some low confident probes and subset the samples to be discovery cohort only.

```{r, message=FALSE}
library(feather)
library(SummarizedExperiment)
library(tidyverse)
```

Read the beta values per probe
```{r}
meth_tbl <- read_feather('../processed_data/cptac3_gbm_epic_meth_array.batch1-9_complete.feather')
```

Read EPIC methylation array hg38 manifest and annotaiton from <https://zwdzwd.github.io/InfiniumAnnotation>
```{r}
# Details about the probe design
EPIC_manifest <- readRDS('../external_data/EPIC.hg38.manifest.rds')
# Annotation
EPIC_anno <- readRDS('../external_data/EPIC.hg38.manifest.gencode.v22.rds')
```


## Add sample annotation
```{r}
sample_anno <- read_tsv('../../annotations/Batches1through9_samples_attribute.with_disease_code.tsv.gz')
cohort_tbl <- read_tsv('../../201907_locate_data/tracked_results/cptac3_gbm_seq_status.tsv')
```

```{r}
coldata_df <- sample_anno %>%
    filter(SAMPLE_ID %in% colnames(meth_tbl)[-1]) %>%
    mutate(
        is_tumor = IS_TUMOR == 'Y',
        cancer_type = 'GBM',
        primary_tumor_location = factor(PRIMARY_TUMOR_LOCATION)
    ) %>%
    select(
        sample=SAMPLE_ID,
        case=SUBJECT_ID,
        cancer_type,
        is_tumor,
        primary_tumor_location
    ) %>%
    filter(case %in% cohort_tbl$case) %>%
    arrange(case)
coldata_df %>% head()
```

C3N-01850 has two arrays:
```{r}
coldata_df %>% 
    count(case, sort = TRUE) %>%
    filter(n > 1)

coldata_df %>%
    filter(case == 'C3N-01850')
```

A quick correlation between the two arrays
```{r}
meth_tbl.C3N_01850 <-  meth_tbl %>%
    select(CPT0125260005, CPT0125270009)
```

```{r}
cor(meth_tbl.C3N_01850$CPT0125260005, meth_tbl.C3N_01850$CPT0125270009,
    use = 'complete.obs')
cor(meth_tbl.C3N_01850$CPT0125260005, meth_tbl.C3N_01850$CPT0125270009,
    use = 'complete.obs', method = 'spearman')
```

```{r, fig.width=7, fig.height=5}
ggplot(meth_tbl.C3N_01850, aes(x = CPT0125260005, y = CPT0125270009)) + 
    geom_bin2d(bins = 50) + 
    coord_fixed()
```

Remove the old array for the sample with duplicates
```{r}
coldata_df <- coldata_df %>%
    filter(sample != 'CPT0125260005') %>%
    column_to_rownames('case')

coldata_df %>% head()
```


## Use the masked probes
Number of probes before filtering
```{r}
nrow(meth_tbl)
```

Select only probes not masked
```{r}
selected_probes <- intersect(meth_tbl$index, EPIC_manifest %>% subset(!MASK_general) %>% names())
length(selected_probes)
```

Convert the methylation array to a matrix
```{r}
meth_mat <- meth_tbl %>%
    filter(index %in% selected_probes) %>%
    column_to_rownames('index') %>% 
    as.matrix()

meth_mat <- meth_mat[, coldata_df$sample]
colnames(meth_mat) <- rownames(coldata_df)
dim(meth_mat)
```

```{r}
selected_anno_gr <- EPIC_anno[selected_probes, ]
```


## Create a SummarizedExperiment object
```{r}
meth_se <- SummarizedExperiment(
    rowRanges = selected_anno_gr,
    colData = coldata_df,
    assays = list(beta_val=meth_mat),
    metadata = list(
        cohort = 'CPTAC3 GBM adhoc cohort (batch 1 to 9; complete)',
        description = 'Beta values of Illumina EPIC DNA methylation microarry',
        pipeline = str_c(
            'Based on the hg38 aligned data from Ding Lab CPTAC methylation pipeline v1.1 ',
            'at https://github.com/ding-lab/cptac_methylation. ',
            'Probes were further filtered, keeping MASK_general == FALSE probes based on the annotation ',
            'from http://zwdzwd.io/InfiniumAnnotation/20180909/EPIC/EPIC.hg38.manifest.rds. ',
            'More information about the annotation ',
            'at https://zwdzwd.github.io/InfiniumAnnotation (DOI: 10.1093/nar/gkw967)'
        ),
        annotation = str_c(
            'Based on GENCODE v22 ',
            'from http://zwdzwd.io/InfiniumAnnotation/20180909/EPIC/EPIC.hg38.manifest.gencode.v22.rds'
        ),
        contact = 'Wen-Wei Liang <liang.w@wustl.edu>, Liang-Bo Wang <liang-bo.wang@wustl.edu>'
    )
)
```

```{r}
saveRDS(meth_se, '../processed_data/gbm_dna_meth.batch1-9_complete.rds')
```

```{r}
coldata_df %>%
    rownames() %>%
    write_lines('../processed_data/cases_with_dna_meth.list')
```


